A coherent and comprehensive dating system for stars in the Milky Way

### 2017-05-11 paper/ms.tex
copying across words written for the ADAP

@@

@@

We explore the parameter space via Gibbs sampling.
Gibbs sampling is a variation of MCMC that makes use of conditional
independence within the problem to reduce dimensionality.
It is often used in hierarchical inference problems such as this, where one
has a set of global parameters ($a$, $b$, $n$, $\alpha$ \& $\beta$) and sets
of per-object parameters ($\{A\}$, $\{M\}$, $\{[Fe/H]\}$, $\{D\}$, $\{A_v\}$).
Instead of updating all parameters simultaneously as in Metropolis,
Metropolis-Hastings, affine invariant \citep{Goodman2010}, nested, Hamiltonian,
and many other sampling algorithms, in Gibbs sampling, subsets of parameters
are updated.
For example, the gyrochronology parameters will be updated, then the kinematic
parameters, then the isochronal parameters for star 1, then star 2, etc.
Since the number of parameters is $N_{gyro} + N_{kinematic} + N_{iso} \times
N_{stars}$, the number of parameters grows rapidly as stars are added.

>>>>

The age-velocity dispersion relations are relatively well studied and based on
simple physics.
There is evidence to suggest that stars in the Milky Way form in the thin disc
of the galaxy with relatively small vertical velocities and angular momenta
\racomment{add citations}.
As time passes these stars are scattered via close encounters with other
stars and interactions with galactic spiral arms.
The more time passes, the more scattering events, and stars slowly accumulate
angular momentum in the vertical direction, known as vertical action.
This results in a slow heating of stellar orbits.
Older stars can be identified in Gaia DR1 by integrating their orbits in the
potential of the Milky Way to convert their proper motions, positions and
parallaxes into vertical actions.
It is the {\it dispersion} in vertical action that truly traces age, far
better than vertical action itself.
However for single stars only the individual's vertical action is available.
Nonetheless, $J_z^2$ increases over time for any given star, and is a weak
tracer of age.
When studying populations of stars, the vertical action dispersion can be used
to infer their age relatively precisely (to within 2 Gyr
\racomment{Ruth, check this}).
However for single stars vertical action can only constrain ages to within 4
Gyr.

@@

{\tt chronometer} extends isochrone fitting by simultaneously fitting an
age-rotation relation and a age-velocity dispersion relation to stars with
photometric colors, rotation periods and Gaia proper motions.
It combines the information available from these three dating techniques, thus
providing more accurate ages than any one of the techniques alone.

The accuracy of ages calculated with {\tt chronometer} still relies on the
rotational and kinematic age relations being correct.
As discussed above, the performance of the age-rotation relations for old
stars has been called into question in the last few years \citep{Angus2015,
Vansaders2016, Metcalfe2016}.
A simple straight-line model for rotational age does not reproduce the ages
predicted by asteroseismology for stars older than the Sun.
This phenomenon is attributed to a transitioning magnetic dynamo at a critical
Rossby number, $Ro$, of 2.6, the solar value \citep{Vansaders2016}.
As rotation periods slow, $Ro$ decreases until it hits the critical value and
magnetic braking switches off: stars maintain their rotation period from that
point onward.
This restricts the applicability of gyrochronology to young, rapidly rotating
stars.
This hypothesis is supported by the existing data, however these data are
sparse~---~the analyses demonstrating the discrepancies were conducted on the
small number of main sequence, Solar-like oscillators observed by Kepler in
short cadence mode with detectable rotation periods: a sample size of around
20 \racomment{Ruth - check this number}.
% Furthermore, asteroseismology favors stars with low surface gravities since
% they show the largest amplitudes of oscillation and the majority of stars in
% this sample were slightly evolved.
A larger sample of old main sequence stars with precise ages is required to
confirm and further characterize the Rossby saturation mechanism introduced by
\citep{Vansaders2016} which is currently supported by the Kepler data.
However, old main sequence stars are difficult to age-date with any other
method than asteroseismology.
We propose use the age gradients in the kinematic properties of stars to
confirm the Rossby saturation effect.

A simple test will allow us to distinguish between two magnetic braking
scenarios in MS stars: the continuous scenario in which stars begin magnetic
braking at zero-age main sequence and continue until main sequence turn off,
and the transition scenario in which stars stop braking at a critical Rossby
number.
This test involves computing rotation-based ages for stars observed by both
Kepler and Gaia.
Each star will have ages computed using the continuous scenario and the
transition scenario.
The vertical action dispersion will then be calculated for groups of stars
with similar rotational age, producing two vertical action-age relations.
The vertical action dispersion will also be calculated for red giants observed
by APOGEE with ages from the Cannon \citep{Ness2015}, observed by Gaia.
This red giant vertical action-age relation will serve as our "ground truth"
since these ages are relatively robust and well calibrated with asteroseismic
giants.
We will compare the both the $\sigma_{J_z}$-age relation for the continuous
scenario and the $\sigma_{J_z}$-age relation for the transition scenario with
the red giant $\sigma_{J_z}$-age relation.
This test will reveal whether continuous magnetic braking or transitional
magnetic braking provides a better description of the data.

This test will allow us to select an appropriate age rotation relation to be
used in {\tt chronometer}, our new age-dating method for MS stars.
{\tt chronometer} combines the following dating methods: gyrochronology,
isochrone fitting, asteroseismology and galactic kinematics into one model by
simply summing the log-likelihoods of the three different models.

@@

We explore the parameter space via Gibbs sampling.
Gibbs sampling is a variation of MCMC that makes use of conditional
independence within the problem to reduce dimensionality.
It is often used in hierarchical inference problems such as this, where one
has a set of global parameters ($a$, $b$, $n$, $\alpha$ \& $\beta$) and sets
of per-object parameters ($\{A\}$, $\{M\}$, $\{[Fe/H]\}$, $\{D\}$, $\{A_v\}$).
Instead of updating all parameters simultaneously as in Metropolis,
Metropolis-Hastings, affine invariant \citep{Goodman2010}, nested, Hamiltonian,
and many other sampling algorithms, in Gibbs sampling, subsets of parameters
are updated at a time.
For example, the gyrochronology parameters will be updated, then the kinematic
parameters, then the isochronal parameters for star 1, then star 2, etc.
Since the number of parameters is $N_{gyro} + N_{kinematic} + N_{iso} \times
N_{stars}$, the number of parameters grows rapidly as stars are added.
